Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2021-03-23 manuscriptPGPEsuppl.tex
Removed 300K MacRog POPE data, because exp data is at 310K, for which there is another data set.
Removed 313K Gromos-CKP POPE data, because exp data is at 310K, for which there is another data set.
Update text

The quality of PE and PG headgroup conformational ensembles in different simulations against NMR experiments is evaluated in figures \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG} using C-H bond order parameters as in our previous studies for PC and PS lipids \cite{botan15,antila19}. Conclusions are the same for all lipids: None of the force fields correctly captures the lipid headgroup conformational ensembles, but CHARMM36 gives results closest to experiments.

@@

Headgroup order parameters of PC lipids are unchanged upon addition
of zwitterionic lipids or cholesterol in experiments, but increase
upon addition of negatively charged PG or PS lipids because
headgroup dipole tilts more parallel to the membrane plane after incorporation
of negative charges into the membrane~\cite{seelig87, scherer87,antila18}.
The response of PC headgroup order parameters to the addition of PE or PG lipids
from different simulations is compared with experiments in figure~\ref{HGorderparametersPCvsPEPG}.
None of the simulations reproduce neither the experimentally observed increase in PC headgroup order parameters
with increasing amount of PG nor the related tilting of the headgroup more parallel with the membrane.
Similar observations in our previous work for PS lipids were explained by the overestimated counterion
binding affinity that neturalizes the effect of added negative charge \cite{antila19}.
All simulations except Berger-OPLS predict tilting of P-N headgroup outwards from the membrane and
decrease of PC headgroup order parameters upon addition of PE lipids.
These results are not in line with experiments where the PC headgroup order parameters are not affected by zwitterionic lipids \cite{scherer87}.
The good performance of Berger-OPLS simulations in here is surprising because
headgroup conformational enemble is not very close to experiments in this model and
the response of headgroup order parameters
to cholesterol was significantly overestimated by the Berger/H{\"o}ltje force field in our previous work \cite{botan15}.

@@

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{./Figs/DIHEDRALS.eps}
  \caption{\label{DIHdists}
    Heavy atom dihedral angle distributions from CHARMM36 simulations that correctly capture the
    order parameter differences between the force fields.
  }
\end{figure}

\clearpage
\section{Changes in headgroup conformations upon addition of charged surfactants or CaCl$_2$}

@@

      POPE  & CHARMM36 \cite{klauda10}           & 0.11   & 500       & 25000 & 50  &  310  & 500 & 100 & \cite{POPEcharmm150mMNaCl} \\

@@

      POPE  & Slipids / {\AA}qvist \cite{jambeck12b,aqvist90}  & 0.11 & 500 & 25000 & 50  &  310  & 500 & 100 & \cite{POPEslipids150mMNaCl} \\

@@

      POPE  & GROMOS-CKP    \cite{??}      &0.11 & 500	& 25000 &50   & 310  & 500 & 100 & \cite{gromosCKPpopeT310150mMNaCl} \\

@@

      POPE             & LIPID17 \cite{gould18} & 0.11   & 500 & 25000 & 50  &  310  & 500 & 100 & \cite{POPElipid17150mMNaCl} \\

@@

      \begin{table*}[htb]
  %\begin{sidewaystable*}[!p]
  \centering
  \caption{List of MD simulations with PG lipids.
    % The salt concentrations calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
    % these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
    % The lipid force fields named as in our previous work~\cite{botan15}.
  }\label{systemsPG}
  \begin{minipage}[t]{\textwidth}
  \resizebox{\columnwidth}{!}{
    \begin{tabular}{l c c r r r r r r c c}
      %\hline
      % some footnotes are not visible in typeset-MS (pdf)
      lipid/counter-ions & force field for lipids / ions & NaCl (M) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
      \hline
      POPG/K$^+$  & CHARMM36 \cite{venable13}       &0         & 118& 4110   &0    & 298  & 100 & 100 & \cite{CHARMM36popg}  \\
      POPG             & CHARMM36 \cite{venable13}        & 0.11           & 500 & 25000 & 49  &  310  & 500 & 100 & \cite{POPGcharmm150mMNaCl} \\
      POPG             & CHARMM36 \cite{venable13}        & 0              & 500 & 25000 & 0   &  310  & 500 & 100 & \cite{POPGcharmm} \\
      \hline
      POPG/Na$^+$  & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}    &0         & 288 	& 10664   &0     & 298  & 250 & 100 & \cite{slipidsPOPGfiles} \\
      DPPG/Na$^+$  & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}    &0         & 288 	& 11232  &0     & 314  & 200 & 100 & \cite{slipidsDPPGfiles} \\
      DPPG/Na$^+$  & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}    &0         & 288 	& 11232   &0     & 298  & 400 & 100 & \cite{slipidsDPPGfilesT298K} \\
      POPG         & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}    & 0    & 500 & 25000 & 0  &  310  & 500 & 100 & \cite{POPGslipids} \\
      POPG         & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}    & 0.11    & 500 & 25000 & 49  &  310  & 500 & 100 & \cite{POPGslipids150mMNaCl} \\
      \hline
      POPG             & LIPID17 / Dang \cite{gould18,smith94,dang06}         & 0              & 500 & 25000 & 0   &  310  & 500 & 100 & \cite{POPGlipid17} \\
      POPG             & LIPID17 \cite{gould18}         & 0.11           & 500 & 25000 & 49  &  310  & 500 & 100 & \cite{POPGlipid17150mMNaCl} \\
      \hline
      POPG             & GROMOS-CKP \cite{??}         & 0              & 500 & 25000 & 0  &  310  & 500 & 100 & \cite{POPGgromosCKP} \\
      POPG             & GROMOS-CKP \cite{??}         & 0.11           & 500 & 25000 & 49 &  310  & 500 & 100 & \cite{POPGgromosCKP150mMNaCl} \\
    \end{tabular}
    }
  \end{minipage}
  \todo{Citation and ion model for GROMOS-CKP?}
\end{table*}
% \end{sidewaystable*}

@@

\noindent {\it POPE with additional NaCl} \todo{Simulation details by A. Peon.}

@@

\noindent {\it POPG with additional NaCl} \todo{Simulation details by A. Peon.}

@@

@@

\noindent {\it POPE with additional NaCl} \todo{Simulation details by A. Peon.
  I have assumed that ion parameters are default Slipids, i.e., {\AA}qvist, please correct if this is not true.
}

@@

\noindent {\it POPG with additional NaCl} \todo{Simulation details by A. Peon.
  I have assumed that ion parameters are default Slipids, i.e., {\AA}qvist, please correct if this is not true.}

@@

\noindent {\it POPC:POPG mixture with additional NaCl} \todo{Simulation details by A. Peon.
  I have assumed that ion parameters are default Slipids, i.e., {\AA}qvist, please correct if this is not true.}

>>>>

The quality of PE and PG headgroup conformational ensembles in different simulations against NMR experiments is evaluated in figures \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG} using C-H bond order parameters as in our previous studies for PC and PS lipids \cite{botan15,antila19}. Conclusions are the same for all lipids: None of the force fields correctly captures the lipid headgroup conformational ensembles, but CHARMM36 gives results closest to experiments. Most importantly for this work, the CHARMM36 captures the distinct headgroup order parameters for PG and PS lipids observed in NMR experiments (Figs. \ref{HGorderParameters} and \ref{structures} in the main text).

@@

Headgroup order parameters of PC lipids are unchanged upon addition
of zwitterionic lipids or cholesterol in experiments, but increase
upon addition of negatively charged PG or PS lipids because
headgroup dipole tilts more parallel to the membrane plane after incorporation
of negative charges into the membrane~\cite{seelig87, scherer87,antila18}.
The response of PC headgroup order parameters to the addition of PE or PG lipids
from different simulations is compared with experiments in figure~\ref{HGorderparametersPCvsPEPG}.
None of the simulations reproduce neither the experimentally observed increase in PC headgroup order parameters
with increasing amount of PG nor the related tilting of the headgroup more parallel with the membrane.
Similar observations in our previous work for PS lipids were explained by the overestimated counterion
binding affinity that neutralizes the effect of added negative charge \cite{antila19}.
All simulations except Berger-OPLS predict tilting of P-N headgroup outwards from the membrane and
decrease of PC headgroup order parameters upon addition of PE lipids.
These results are not in line with experiments where the PC headgroup order parameters are not affected by zwitterionic lipids \cite{scherer87}.
The good performance of Berger-OPLS simulations is surprising here because
headgroup conformational enemble is not very close to experiments in this model and
the response of headgroup order parameters
to cholesterol was significantly overestimated by the Berger/H{\"o}ltje force field in our previous work \cite{botan15}.

@@

\subsection{Dihedral angles and relative energies of PC, PE, PG and PS headgroups}

We estimated the relative energies of each dihedral angle value
with respect to the most probable value (lowest energy) 
from the inverse Boltzmann formula $\Delta E(\theta) = -kT \left[\ln\left[p(\theta)\right]-\ln\left[p(\theta_0)\right] \right]$,
where $p(\theta)$ is the dihedral angle distribution and $\theta_0$ is the most probable angle from MD simulation.
The dihedral angle distributions are shown in Fig. \ref{DIHdists}
and relative energies in Fig. \ref{structures} in the main text.

\begin{figure}[]
  \centering
  \includegraphics[width=7.0cm]{./Figs/DIHEDRALS.eps}
  \caption{\label{DIHdists}
    Heavy atom dihedral angle distributions from CHARMM36 simulations that correctly capture the
    order parameter differences between the force fields.
  }
\end{figure}

\clearpage
\subsection{Changes in headgroup conformations upon addition of charged surfactants or CaCl$_2$}

@@

%      POPE  & CHARMM36 \cite{klauda10}           & 0.11   & 500       & 25000 & 50  &  310  & 500 & 100 & \cite{POPEcharmm150mMNaCl} \\

@@

%      POPE  & Slipids / {\AA}qvist \cite{jambeck12b,aqvist90}  & 0.11 & 500 & 25000 & 50  &  310  & 500 & 100 & \cite{POPEslipids150mMNaCl} \\

@@

%      POPE  & GROMOS-CKP    \cite{??}      &0.11 & 500	& 25000 &50   & 310  & 500 & 100 & \cite{gromosCKPpopeT310150mMNaCl} \\

@@

%      POPE             & LIPID17 \cite{gould18} & 0.11   & 500 & 25000 & 50  &  310  & 500 & 100 & \cite{POPElipid17150mMNaCl} \\

@@

      \begin{table*}[htb]
  %\begin{sidewaystable*}[!p]
  \centering
  \caption{List of MD simulations with PG lipids.
    % The salt concentrations calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
    % these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
    % The lipid force fields named as in our previous work~\cite{botan15}.
  }\label{systemsPG}
  \begin{minipage}[t]{\textwidth}
  \resizebox{\columnwidth}{!}{
    \begin{tabular}{l c c r r r r r r c c}
      %\hline
      % some footnotes are not visible in typeset-MS (pdf)
      lipid/counter-ions & force field for lipids / ions & NaCl (M) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
      \hline
      POPG/K$^+$  & CHARMM36 \cite{venable13}       &0         & 118& 4110   &0    & 298  & 100 & 100 & \cite{CHARMM36popg}  \\
%      POPG             & CHARMM36 \cite{venable13}        & 0.11           & 500 & 25000 & 49  &  310  & 500 & 100 & \cite{POPGcharmm150mMNaCl} \\
      POPG             & CHARMM36 \cite{venable13}        & 0              & 500 & 25000 & 0   &  310  & 500 & 100 & \cite{POPGcharmm} \\
      \hline
      POPG/Na$^+$  & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}    &0         & 288 	& 10664   &0     & 298  & 250 & 100 & \cite{slipidsPOPGfiles} \\
      DPPG/Na$^+$  & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}    &0         & 288 	& 11232  &0     & 314  & 200 & 100 & \cite{slipidsDPPGfiles} \\
      DPPG/Na$^+$  & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}    &0         & 288 	& 11232   &0     & 298  & 400 & 100 & \cite{slipidsDPPGfilesT298K} \\
      POPG         & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}    & 0    & 500 & 25000 & 0  &  310  & 500 & 100 & \cite{POPGslipids} \\
%      POPG         & Slipids / {\AA}qvist \cite{jambeck13,aqvist90}    & 0.11    & 500 & 25000 & 49  &  310  & 500 & 100 & \cite{POPGslipids150mMNaCl} \\
      \hline
      POPG             & LIPID17 / Dang \cite{gould18,smith94,dang06}         & 0              & 500 & 25000 & 0   &  310  & 500 & 100 & \cite{POPGlipid17} \\
%      POPG             & LIPID17 \cite{gould18}         & 0.11           & 500 & 25000 & 49  &  310  & 500 & 100 & \cite{POPGlipid17150mMNaCl} \\
      \hline
      POPG             & GROMOS-CKP \cite{??}         & 0              & 500 & 25000 & 0  &  310  & 500 & 100 & \cite{POPGgromosCKP} \\
%      POPG             & GROMOS-CKP \cite{??}         & 0.11           & 500 & 25000 & 49 &  310  & 500 & 100 & \cite{POPGgromosCKP150mMNaCl} \\
    \end{tabular}
    }
  \end{minipage}
  \todo{Citation and ion model for GROMOS-CKP?}
\end{table*}
% \end{sidewaystable*}

@@

%\noindent {\it POPE with additional NaCl} \todo{Simulation details by A. Peon.}

@@

%\noindent {\it POPG with additional NaCl} \todo{Simulation details by A. Peon.}

@@

\noindent {\it POPC and cationic surfactant (dihexadecyldimethylammonium) mixture}
Intial structures were taken from similar previously published \cite{melcr18} simulations with Amber lipid14 force field,
which are available from Ref. \citenum{POPClipid14T313K,POPClipid1410perCATsurfT313K,POPClipid1420perCATsurfT313K,POPClipid1430perCATsurfT313K,POPClipid1442perCATsurfT313K,POPClipid1450perCATsurfT313K}.
Default simulations parameters and force field files from CHARMM-GUI \cite{lee16} were used, except for dihexadecyldimethylammonium
for which the atom types and partial charges of Amber lipid14 parameters from previous work \cite{melcr18} were modified to correspond Charmm36
force field. 
Systems contained 50 POPC molecules, 3983 water molecules, and 12, 30, 44, or 88 dihexadecyldimethylammonium molecules.
Chloride ions were used as counterions for dihexadecyldimethylammonium.
Reference system without cationic surfactants contained 200 POPC and 9000 water molecules.
Systems were simulated 200 ns (the first 20 ns was discarded as an equilibration period)
using Gromacs 5 \cite{abraham2015gromacs} at the temperature of 313 K.
All simulation files are available from Refs. \citenum{charmmPOPC313K,CHARMM36cationicSURF}.

@@

%\noindent {\it POPE with additional NaCl} \todo{Simulation details by A. Peon.
%  I have assumed that ion parameters are default Slipids, i.e., {\AA}qvist, please correct if this is not true.
%}

@@

%\noindent {\it POPG with additional NaCl} \todo{Simulation details by A. Peon.
%  I have assumed that ion parameters are default Slipids, i.e., {\AA}qvist, please correct if this is not true.}

@@

%\noindent {\it POPC:POPG mixture with additional NaCl} \todo{Simulation details by A. Peon.
%  I have assumed that ion parameters are default Slipids, i.e., {\AA}qvist, please correct if this is not true.}
